Epistemic Semantics in Guarded String Models

### 2020-07-21 tex/draft.tex
Edits in Section 3 of draft.tex

\begin{example}  \label{raise1}
  Let $W$ be a set of guarded strings over  events P and primitive tests B, a relation ${R}$ be a relation
  on $P$. The corresponding relation $\hat{R}$ on $W$ holds between a guarded string
\[
\alpha_0e_0 ... e_n\alpha_{n+1} 
\]
in $W$ and a guarded string  $q$ iff $q$ is an element of $W$ and is of the form

@@

Each term in the language is interpreted as a set of guarded strings.  Where $p$ is a
term, $\sem{p}^{\test,\event,\varphi,\zeta}$ is the interpretation of the term.  The
superscript captures dependence on an Epik specification.

interprets a term $p$ as a set of
guarded strings consistent with state constraints $\varphi$ and effect
relations ${\mathcal{E}^\varphi}$. A bare event $e$ is interpreted as
the set $\hat{\mathcal E}^\varphi(e)$. The constant $0$ is interpreted
as the empty set.  The constant $1$ is interpreted as the set
$\mathcal A^\varphi_{\mathsf B}$, i.e. $\{h\bar t, \bar h t\}$ in the
coin example. The operation $p + q$ is interpreted as the union of the
interpretations of $p$ and $q$. The operation $p;q$ interpreted as the
fusion product of the interpretations of $p$ and $q$ raised to sets:
$X \diamond Y = \{ x \diamond y \mid x \in X, y \in Y, x\diamond
y~\textit{defined}\}$.  The operation $p*$ is Kleene star, which is
interpreted to be least upper bound of the iterated fusion product of
the denotation of $p$ with itself. Subsets of $1$ indicated by
propositions $\psi$ are also elements of $K$, and these form the
Boolean algebra of tests, denoting all atoms admitted by the
constraint $\varphi$ and the boolean test $\psi$.

The unary epistemic alternative operation $\Diamond_a$ for each agent
$a$, is interpreted using using Kripke semantics, as pre-image
relative to a fixed relation $\hat{R_a}$ between guarded strings,
$\Diamond_a~p = \{u \mid \exists v. v \in \sem{
p}^{\phi, \mathcal E} \wedge u \mathrel{\hat{R_a}} v\}$.  Here $u$
and $v$ are guarded strings, while $x$ is an element of $K$.

@@

\newcommand{\epikspec}{\test,\event,\varphi,\zeta}

\newcommand{\AphiB}{\mathcal A^\varphi_{\mathsf B}}

\begin{figure}
  \[\begin{array}{l >{\triangleq}c l}
  \sem{ 0 }^{\epikspec} && \emptyset \\  
  \sem{ 1 }^{\epikspec} && \mathcal A^\varphi_{\mathsf B}\\
  \sem{ a }^{\epikspec} && \mathcal A^{a\varphi}_{\mathsf B}\\
  \sem{ \bar{\sigma} }^{\epikspec} && \AphiB \setminus \sem{\sigma}^{\epikspec}\\
  \sem{ e }^{\epikspec}&&  \setabs{\alpha e \beta}{\alpha \zeta_e \beta} \\
  \sem{{p+q}}^{\epikspec} && \sem{{p}}^{\epikspec} \cup \sem{ q }^{\epikspec}\\
  \sem{{pq}}^{\epikspec} && \setabs{x \diamond y}{ \begin{array}{l}x \in \sem{p}^{\epikspec} \\ y \in \sem{q}^{\epikspec}\\\mbox{$x \diamond y$ is defined}\end{array}} \\
  \sem{ p^*}^{\varphi,{\mathcal{E}}} && \displaystyle\bigcup_n {\underbrace{\sem{ p }^{\varphi,{\mathcal{E}}}\diamond \ldots \diamond \sem{ p }^{\varphi,{\mathcal{E}}}}_{n}} \\
  \sem{ \neg p }^{\varphi,{\mathcal{E}}} && \sem{ \_^*}^{\varphi,{\mathcal{E}}} \setminus \sem{ p }^{\varphi,{\mathcal{E}}} \\
  \sem{ \Diamond_a~p } && \{x \mid x \mathbin{\hat{R_a}} y, y \in \sem{ p }^{\varphi,\mathcal E}\}

  \end{array}
  \]
  \caption{Interpretation of Epik terms as sets of guarded strings}
  \label{fig:interp2}
\end{figure}

@@

The complement operation ${p}^c$ is complement at the level of sets of
guarded strings, defined to every set the universe $\_^*$ except those
denoted by $p$. Note that $\sem{ 0^c
}^{\varphi,{\mathcal{E}}} =
\sem{\_^*}^{\varphi,{\mathcal{E}}}$.

Summing up, given an Epik program with $n$ agents, we construct state
constraints $\varphi$, effect relations ${\mathcal{E}}$, and a concrete guarded
string model for each term using $\sem{ -
}^{\varphi,{\mathcal{E}}}$.
%% $\tuple{K,+,;,*,\bar{},0,1,{}^c,\Diamond_{a_1},...,\Diamond_{a_n}}$.
 $0^c$ is the set of worlds, and it may be countably infinite.
$\Diamond_{a_i}$ is an epistemic modality for the $i$th agent.  Or
referring to the Kripke relations $\hat{R}_i$, the construction
defines a multi-agent Kripke frame
$\tuple{0^c,\hat{R}_1,...,\hat{R}_n}$ (usually a countable one) from
an Epik specification. The frame consists of a set of worlds, and an
epistemic-alternative relation for each agent.  These models are used
as a target for natural-language interpretation in Section 5 and
Section 6, where we obtain semantic values such as \sem{Amy knows that
  it's heads and Bob knows that Amy knows whether it is heads or
  tails, and does not know that it's heads} as elements of
$K$. Concretely the propositions are sets of guarded strings (usually
countable ones), construed as sets of worlds as they figure in
possible worlds semantics for natural language.

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics 
\citep{kaplan1994regular,beesley2003finite}.
Current implementations allow for the
definition of functions with the status of defined operators on regular sets and
relations
\citep{hulden2009foma,karttunen2010update}.
Such definitions are used to define an embedding of epistemic KAT in a
string algebra. The methodology follows Section 2 closely.  Let ${\cal K}$ be an
epistemic algebra as described in Section 2.  A given element of ${\cal K}$ is
represented in the the string algebra by the very same set of strings, i.e. by a set
of strings that have the form of a sequence of bare event symbols, with interleaved
Boolean vectors. Product in the KAT can not be modeled as concatenation in the string
algebra, because this would not enforce identity of states, and would result in
lengthening Boolean vectors at the concatenation point.  Instead, KAT product and KAT
Kleene star are defined operations in the string algebra, see Figure \ref{fstfigure1}.  The
operations concatenate in the string algebra, delete strings with non-matching tests,
and then delete the second of two tests create a well-formed guarded string.

>>>>

\begin{example}  \label{raise1}
  Let $W$ be a set of guarded strings over  events P and primitive tests B, and ${R}$ be a relation
  on $P$. The corresponding relation $\hat{R}$ on $W$ holds between a guarded string
\[
\alpha_0e_0 ... e_n\alpha_{n+1} 
\]
in $W$ and a guarded string  $q$ iff $q$ is an element of $W$ and is of the form

@@

\newcommand{\epikspec}{\test,\event,\varphi,\zeta}
\newcommand{\AphiB}{\mathcal A^\varphi_{\mathsf B}}

Each term in the language is interpreted as a set of guarded strings.  Where $p$ is a
term, $\sem{p}^{\epikspec}$ is the interpretation of the term, where superscript
captures dependence on an Epik specification.  Figure \ref{fig:interp2} defines the
interpretation of terms recursively.  The interpretation $\sem{ 1 }^{\epikspec}$ of
the multiplicative identity is the set of atoms that satisfy the state constraint
$\varphi$.  Where $b$ is a primitive Boolean, $\sem{ b }^{\epikspec}$ is the set of
atoms that satisfy the state constraint and where $b$ is true.  Where $e$ is a
primitive event, $\sem{ e }^{\epikspec}$ is the set of guarded strings that have the
form of $e$ flanked by compatible atoms, as determined by the event formula
$\zeta_e$. The product $pq$ of terms $p$ and $q$ is interpreted with fusion product
raised to sets of guarded strings.
Kleene star is interpreted as the union of iterated products. $p^n$ is the $n$-times product
of $p$ with itself, with $p^0 = 1$.
Propositional complement is complement relative to the set of worlds.
The epistemic formula $\Diamond_a p$ is interpreted with Kripke
semantics for epistemic modality, as the pre-image of the
complement proposition under the relation
$\hat{R}_a$.  Here $\hat{R}_a$ is the world-alternative for agent $a$,
as defined in \ex{raise1}.

@@

\begin{figure}
  \(\begin{array}{l >{\triangleq}c l}
  \sem{ 0 }^{\epikspec} && \emptyset \\  
  \sem{ 1 }^{\epikspec} && \mathcal A^\varphi_{\mathsf B}\\
  \sem{ b }^{\epikspec} && \mathcal A^{b\varphi}_{\mathsf B}\\
  \sem{ \bar{\sigma} }^{\epikspec} && \AphiB \setminus \sem{\sigma}^{\epikspec}\\
  \sem{ e }^{\epikspec}&&  \setabs{\alpha e \beta}{\alpha \zeta_e \beta} \\
  \sem{{p+q}}^{\epikspec} && \sem{{p}}^{\epikspec} \cup \sem{ q }^{\epikspec}\\
  \sem{{pq}}^{\epikspec} && \setabs{x \diamond y}{ \begin{array}{l}x \in \sem{p}^{\epikspec} \\ y \in \sem{q}^{\epikspec}\\\mbox{$x \diamond y$ is defined}\end{array}} \\
  \sem{p\!*\!}^{\epikspec} && \displaystyle\bigcup_{n \geq 0} {\sem{ p^n }^{\epikspec}} \\
  \sem{ \neg p }^{\epikspec} && \sem{\bullet\!*}^{\epikspec} \setminus \sem{ p }^{\epikspec} \\
  \end{array}
  \)

\hspace{2mm}\(\sem{ \Diamond_a\,p } \hspace{4mm}{\triangleq} \hspace{4mm}\{x \mid \exists y . x{\hat{R}_a}y \wedge y \in \sem{ p }^{\epikspec}\} \)
  \caption{Interpretation of Epik terms as sets of guarded strings (Version M)}
  \label{fig:interp2}
\end{figure}

@@

Summing up, given an Epik specification, a state constraint $\varphi$, and effect
relation $\zeta_e$ for each primitive are constructed. Given these, each term $p$ (as
defined syntactically in Figure \ref{termfigure}) is interpreted as a set of guarded
strings $\sem{p}^{\epikspec}$ over the Boolean primitives and event primitives in the
specification.  Let $K^{\epikspec}$ be the sets that are interpretations of terms.
Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of epistemic KAT, with
operations as in Figure \ref{fig:interp2} (e.g. the binary operation $+$ is union,
and the unary operation $\Diamond_a$ is pre-image relative to $\hat{R}_a$).  The
construction can also be viewed as providing a concrete $n$-agent Kripke frame
$\tuple{\sem{\bullet *}^{\epikspec}, \hat{R}_1,...,\hat{R}_n}$.\footnote{ The domain
  of the Kripke frame differs from the domain of the guarded string model, because
  the former is the set of worlds, while the latter is the set of propositions.  }
The frame consists of a set of worlds, and an epistemic-alternative relation for each
agent.  These models are used as a target for natural-language interpretation in
Section 5 and Section 6.

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics \citep{kaplan1994regular,beesley2003finite}.  Current
implementations allow for the definition of functions that have the status of defined
operations on regular sets and relations \citep{hulden2009foma,karttunen2010update}.
Such definitions are used here to construct of a model for epistemic KAT inside the
finite state calculus.  The space of worlds is a set of ordinary strings.  Bit
sequences (sequences of 0's and 1's) are used for atoms, and these alternate with
event symbols in the encoding of a world. \ex{gs1fst} gives the encoding of some of
the worlds from the examples.  A string is a finite sequence of symbols, and {\tt 0},
{\tt 1}, {\tt u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are special
multi-character symbols that are found in implementations of the finite state
calculus.

\begin{example}   \label{gs1fst}
  \renewcommand{\arraystretch}{1.2}
Worlds coded as strings
\vspace{1mm}

 \hspace{-13mm} \begin{tabular}[t]{cl}
{\textit{World}} & {\textit{String}} \\
% $v_1$ & $ 0 \,u\,  1 \,u\,  2 \,d\,  1 \,u\,  2$ \\
$v_1$ & {\tt 0 0 u 0 1 u 1 0 d  0 1 u  1 0} \\
$v_2$ & $ 0 \,u\,  1 \,u\,  2 \,u\,  3$ \\   
$w_1$ & $1 0$ \\
$w_2$ & $1 0 \,a_0\, 1 0$ \\  
$w_3$ & $0 1 \,a_0\, 0 1 \,b_0\, 0 1$ \\
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

Operations such as KAT product are defined operations in the finite
state calculus.

of epistemic KAT in a
string algebra. The methodology follows Section 2 closely.  Let ${\cal K}$ be an
epistemic algebra as described in Section 2.  A given element of ${\cal K}$ is
represented in the the string algebra by the very same set of strings, i.e. by a set
of strings that have the form of a sequence of bare event symbols, with interleaved
Boolean vectors. Product in the KAT can not be modeled as concatenation in the string
algebra, because this would not enforce identity of states, and would result in
lengthening Boolean vectors at the concatenation point.  Instead, KAT product and KAT
Kleene star are defined operations in the string algebra, see Figure \ref{fstfigure1}.  The
operations concatenate in the string algebra, delete strings with non-matching tests,
and then delete the second of two tests create a well-formed guarded string.
